---
title: "Repeated_Measures_ANOVA"
author: "Kristopher Kyle"
date: "Last updated 2022-01-29"
output:
  html_document:
    toc: true
    toc_float: true
---
[Back to Homepage](https://kristopherkyle.github.io/IntroQuantALRM/)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Repeated Measures ANOVA
This tutorial will cover the repeated measures analysis of variance (ANOVA) test, which has traditionally been used as the multivariate (or multi-group) alternative to the paired samples t-test. Note that newer, arguably better methods are currently being used as well, namely linear mixed-effects models (these will be covered in the next tutorial). Accordingly, this tutorial will be rather brief. 

### Description of sample data

This data comprises L2 English essays written over a two year period by nine middle-school aged Dutch children studying at an English/Dutch bilingual school in the Netherlands. Essays were collected three times a year (roughly every four months) over two academic years. Included in the dataset are holistic scores for each essay ("Score") and mean length of T-unit (MLT) values. In this tutorial, we will explore the relationship between holistic scores and time spent studying English, with the alternative hypothesis that holistic essay scores will increase as a function of time. For further reference, see Kyle (2016).

```{r}
mydata <- read.csv("data/RM_sample.csv", header = TRUE)
#First, we create a new variable that is the categorical version of Time
mydata$FTime <- factor(mydata$Time)
summary(mydata)
```

### Visualizing the data

First, we can look at the means at each time point.

```{r}
library(ggplot2)
ggplot(data = mydata, aes(x = FTime, y = Score, group = Time)) +
  geom_boxplot()
```

Then, we can get a clearer view by looking at individual trajectories.

```{r}
library(ggplot2)
ggplot(data = mydata, aes(x = FTime, y = Score, group = Participant)) +
  geom_boxplot(aes(group=Time)) +
  geom_line(aes(color=Participant)) +
  geom_point(aes(color=Participant))
```
Sometimes, we get a clearer view of individual trajectories when we use facet wrap:

```{r}
ggplot(data = mydata, aes(x = FTime, y = Score, group = Participant)) +
  geom_line(aes(color=Participant)) +
  geom_point(aes(color=Participant)) +
  facet_wrap(~Participant)
```

### Conducting repeated measures ANOVA

Below, we conduct the repeated measures ANOVA. Note that this is almost the same as an independent one-way ANOVA, but there is one addition. We account for that fact that each participant wrote multiple essays in our sample by adding an "Error" term (in this case, "Participant"). As we can see, there are significant differences between at least two of our groups (p = 7.2e-09).

```{r}
rm_anova <- aov(Score~FTime + Error(Participant), data=mydata)
summary(rm_anova)
```

For this analysis, we will cheat just a bit with the effect size, and use R^2 (which, as we recall is the same as eta squared). As we see below, our R^2 value (and our eta-squared value) indicates a large (R^2 = .581) effect. There may be more appropriate effect sizes to run here, but RM ANOVAs are one of the few analyses that are difficult to run in R. The good news is that there is a reason for this, namely that linear mixed effects models are superior in most cases to RM ANOVAs, and are extensively supported in R.

```{r}
cor(mydata$Time, mydata$Score)^2
```

### Post-hoc analyses
We will have to do a fair bit of work here to get all of our pairwise comparisons. Below is the pairwise contrast between Time 1 and 2. We see that there is a significant difference (p = 0.012) in Score between Time 1 and 2 and that there is a large effect (d = 1.8). Note that if we do multiple pairwise comparisons, we will need to adjust our alpha for the number of tests conducted. Because we are going to look at a better method (linear mixed effects models) soon, we will skip the extra work here.

```{r}
library(dplyr)
library(psych)
#create new df with only Time 1 and 2
Time_12 <- mydata %>%
  filter(FTime == "1" | FTime == "2") %>%
  select(FTime, Score, Participant)
summary(Time_12)

# Conduct paired samples t-test:
t.test(Score~FTime, paired = TRUE, data = Time_12)
# Get effect size
cohen.d(Time_12, "FTime")
```


## Two-Way RM ANOVA

We will now use some sample data to conduct a two way repeated measures ANOVA. In the sample data, we have two groups (A and B) and vocabulary test scores that represent pre, post, and delayed post tests.

```{r}
mydata2 <- read.csv("data/factorial_rm_anova_sample.csv", header = TRUE)
#First, we create a new variable that is the categorical version of Time
mydata2$Participant_ID <- factor(mydata2$Participant_ID)
# then, we will order our tests (pre, post, delayed-post)
mydata2$Time = factor(mydata2$Time, levels = c("Pre","Post","D_Post"))
summary(mydata2)
```

### Visualizing the data

First, we can create boxplots. 

```{r}
library("ggpubr")
ggboxplot(mydata2, x = "Time",
          y = "V_Test", 
          color = "Group")
```

We can also visualize the data per participant:

```{r}
library("ggplot2")
ggplot(mydata2, aes(x=Time, y=V_Test)) +
  geom_boxplot() +
  geom_point(aes(color = Participant_ID), show.legend = FALSE) +
  geom_line(aes(group = Participant_ID, color = Participant_ID), show.legend = FALSE) +
  facet_wrap(~Group)
```

Finally, we will run the analysis (and interpret our results).

```{r}
library(car)
rmf_anova <- aov(V_Test ~ Time * Group + Error(Participant_ID/(Time * Group)), data = mydata2)
summary(rmf_anova)
#Anova(rmf_anova, type = "III")
```

So, as we can see, there is a significant effect for participants within each group (they seem to behave differently), a significant main effect for Time, and a significant interaction between Time and Group. In other words, there are significant differences between the tests, and that there are differences between the groups (somewhere).

### For later...
Because the general consensus has been to use linear mixed effects models, we are not going to go into the post-hoc analyses here. We could do it (e.g., with dplyr and by calculating our own p-value corrections), but it would be fairly burdensome.

See the following tutorials for more information about completing a two-way repeated measures ANOVA: https://datascienceplus.com/two-way-anova-with-repeated-measures/
https://sapa-project.org/blog/2013/06/28/repeated-measures-anova-in-r/


